\ripple: Differentiable and Hardware-Accelerated Waveforms for Gravitational Wave Data Analysis

### 2023-01-20 src/tex/ms.tex
merging adams comments

\definecolor{dfm}{HTML}{6C71C4}
\newcommand{\dfm}[1]{{\color{dfm}[DFM: #1]}}

@@

\newcommand{\JHU}{\affiliation{William H. Miller III Department of Physics and Astronomy, Johns Hopkins University, Baltimore, Maryland 21218, USA}} 
\newcommand{\OKC}{\affiliation{The Oskar Klein Centre, Department of Physics, Stockholm University, AlbaNova, SE-106 91 Stockholm, Sweden}} 
\newcommand{\NORDITA}{\affiliation{Nordic Institute for Theoretical Physics (NORDITA), 106 91 Stockholm, Sweden}}
\newcommand{\UdeM}{\affiliation{Département de Physique, Université de Montréal, 1375 Avenue Thérèse-Lavoie-Roux, Montréal, QC H2V 0B3, Canada}} 
\newcommand{\Mila}{\affiliation{Mila -- Quebec AI Institute, 6666 St-Urbain, \#200, Montreal, QC, H2S 3H1}} 
\newcommand{\CGP}{\affiliation{Center for Gravitational Physics, University of Texas at Austin, Austin, TX 78712, USA}}
\newcommand{\CCA}{\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}}
\newcommand{\CUHK}{\affiliation{Department of Physics, The Chinese University of Hong Kong, Shatin, N.T., Hong Kong}}

@@

% Abstract with filler text
\begin{abstract}
    Since their discovery in 2015, gravitational waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
    As more detections continue, data analysis tasks are becoming increasingly computationally demanding.
    \kw{I would cut everything before out.}
    Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analysis tasks.
    Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
    Moreover, \jax allows for GPU-accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
    We then focus on three applications where efficient and differentiable waveforms are essential.
    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
    In particular, we demonstrate that the typical \textit{match} with numerical relatively waveforms can be improved by more than 10\% without any additional overhead.
    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
    This increased speed makes population forecasting substantially simpler, as discussed extensively in \citet{Iacovelli:2022bbs, Iacovelli:2022mbg}.
    Finally, we show that gradient-based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to traditional Monte Carlo methods.
    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are formulated.

@@

Derivatives are ubiquitously useful throughout data analysis tasks.
For instance, during PE, gradients are used to improve performance of the samplers 
% \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
(e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
Gradients are particularly valuable for high dimensional spaces.
Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain~\citep{Keppel:2013kia}.
\kw{This needs some rewording. For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms. Instead of difficult to have the derivatives, I would rather mention it is clumpsy to get the derivatives, especially given the model complexity is growing.}
Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calculations for gravitational waveforms, with applications to constructing template banks~\citep{Coogan:2022qxs} and computing the Fisher information for forecasting~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.

Automatic differentiation is a family of methods used to compute machine precision derivatives with little computational overhead. 
AD's recent ascendance is primarily driven by its use in machine learning, particularly for derivative computations of neural networks which use gradient descent during training.
The core idea of AD is that any mathematical function can be broken down into a small set of basic operations, each with a known differentiation rule.\footnote{
    Of course, non-differentiable functions exist and care must be taken when treating these special cases.
    }
The full derivative can then be constructed using the chain rule.
There are now a variety of AD implementations, most notably in deep learning frameworks such as \texttt{pytorch}~\citep{pytorch} and \texttt{tensorflow}~\citep{tensorflow2015-whitepaper}.
More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW analysis software precludes its general use.
Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries and its just-in-time (JIT) compiler which allows the same code to run on a variety of hardware accelerators (i.e.~GPUs).

@@

In this paper we argue that differentiable waveforms will be a vital component for the future of GW data analysis.
\AZ{I prefer to describe what is rather than what could be; can we say what elements exist, e.g.~
we present \ripple... which includes \jax implementations of GW antenna patterns, the IMRPhenomD waveform, and...}
In addition, we present \ripple, a small GW \texttt{python} package which can serve as a repository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses. 
The remainder of this paper is structured as follows. 
In Sec.~\ref{sec:waveforms} we discuss the differentiable IMRPhenomD waveform implemented in \ripple and perform some benchmarks to demonstrate its speed and accuracy. 
In Sec.~\ref{sec:applications} we discuss three distinct applications using differentiable waveforms. 
Firstly, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting enabled by a differentiable waveform. 
Secondly, we implement differentiable detector response functions and show that the speed of Fisher matrix calculations can be substantially improved using AD.
Finally, we run an illustrative injection example using Hamiltonian Monte Carlo to demonstrate that the autocorrelation of derivative samplers is substantially smaller than that of traditional MCMCs.
The associated code can be found at \ripple~\citep{ripple}.

@@

A variety of waveform families have been developed to accurately model the GW emission from COs~\citep{Schmidtreview}. 
When the COs are relatively well separated, the dynamics of the system can be well approximated by a post-Newtonian expansion.
However, close to merger, numerical relatively simulations are required to accurately model the binary.
Unfortunately, these numerical simulations are computationally expensive and cannot be run in conjunction with data analysis.
Approximate, phenomenological waveforms have therefore been constructued to enable relatively fast waveform generation at sufficient accuracy.

As mentioned in the previous section, the three major waveform families that have been developed to date are:  EOB,  NRsurrogate, and IMRPhenom.
Note that unlike NRsurrogate, both the EOB and PhenomIMR waveforms are calibrated to numerical relativity waveforms.
EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
IMRPhenom waveforms are therefore ideally suited for AD, espeically using \jax. 
In this paper we focus on the aligned spin circular orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.

@@

For maximum utility, a waveform needs to be fast to evaluate.
Fortunately, the IMRPhenom waveforms are constructed from simple closed-form expressions which are computationally efficient.
Since the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the evaluation of the waveform itself is faster than \ripple. \dfm{I'd soften this statement a little bit. In my own work, I often find that JAX implementations are just as fast or faster than C implementations, so I wouldn't state it that it's obvious that lalsuite should be faster! Where are these benchmarks implemented? Would it be easy for me to try them out?}
Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 2-5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 5-8\,\mathrm{ms}$ for \ripple.\footnote{
    For this benchmark we used $16\,\mathrm{Hz}$ and $1024\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
    In addition, we used a frequency spacing of $0.0125\,\mathrm{Hz}$. 
    We performed this benchmark by evaluating the waveform $5000$ times and taking the average evaluation time.
}
Although slower on a CPU benchmark, a key advantage of using \jax is its native ability to run on a GPU.
Waveform calls can therefore be evaluated in parallel, massively improving performance.
Performing the same benchmark as above on a Google Colab notebook with the freely provided GPU, we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
\kw{We should add some more detail related to this comparison. We should be clear that each individual waveform sample evaluation doesn't get us the performance by putting it on GPU (Jitting helps with that, not the GPU part). The GPU comes into play by allowing parallel evaluation of waveform sample, so for a tiny amount of time/frequency samples, we should expect GPU to lose to a CPU, but on any reasonable amount of time/frequency samples, GPU win a lot.}
Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.

The primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are significantly more stable than finite difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimenison.\footnote{
    Note that although the number of function calls is small, reverse mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
\kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
However, this might change for waveforms in beyond-GR models, \AZ{waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
In these cases} many more parameters \AZ{can be} added.

@@

\subsection{Finetuning Waveform Coefficients}
\label{subsec:coeffs}

@@

Having an accurate waveform model is essential for many data analysis tasks in
GW, such as searching for signals and estimating source parameters given data.
While waveforms generated using numerical relativity simulations are \sout{commonly
regarded as the most accurate,} 
\AZ{in principle the highest fidelity signal model,}
they are too computationally expensive to be used
in any practical data analysis tasks. To create waveform models that can be used
in data analysis, there are a number of groups creating ``approximants'' of the
full numerical relativity waveforms, which are essentially simpler ansatzes that
\sout{are only approximations but} can be evaluated much faster \AZ{and in regions not
covered by numerical simulations, such as extremes of parameter space or earlier 
phases of the binary inspiral}.

@@

Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
coefficients are fitted in segments with amplitude and phase separated.
Furthermore, IMRPhenomD is divided in three fitting segments, inspiral, merger and ringdown.
Each of these segments has their own set of fitting coefficients.
After obtaining the fitting coefficient for individual segments, they are then
``stitched'' together such that they are continuous in the first derivative.
The process of stitching introduces some addition
inaccuracy in the waveform model, as the connections would affect the originally
fitted segments.
The coefficients of the current IMRPhenomD implementation are tuned in sub-sets
of parameters instead of all together. This means the tuning process ignore the
correlation between different sub-sets of parameters, so the best-fitted
solution obtained by fitting sub-sets of parameters may not be the global
optimum compared to fitting all coefficients at once. By jointly
fitting all coefficients, the correlation in fitting individual and connecting
different segments can be accounted for during the optimization. Therefore, there is
potential improvement of the accuracy by fitting all coefficients at once.

@@

The first step in fitting all coefficients is to define a loss function that
measures the goodness-of-fit of the optimization procedure, which we choose it
to be the mismatch between the NR waveform and the approximant waveform.
\begin{equation}
    \label{eq:mismatch}
	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}}(\lambda)),
\end{equation}
where $\lambda$ is a vector of the fitting coefficients, $h_{\mathrm{IMR}}$ is
the waveform generated by the IMRPhenomD and $h_{\mathrm{NR}}$ is the waveform
generated by the numerical relativity simulation. 
Given the loss function, we use gradient descent to find the optimal fitting
coefficients.
\begin{equation}
	\lambda\leftarrow\lambda-\alpha\nabla\mathcal{M}, 
\end{equation}
where $\alpha$ is the learning rate. We set $\alpha$ to be a very small number, i.e. $10^{-6}$, since the initial guess already lies close to the desired optimum. We terminate the program after 30000 steps, in which $\lambda$ has converged to the optimum.  \dfm{We can't possibly know that it has converged to ``the optimum''. I'd sat just leave it at ``We terminate the program after 30000 steps.''}
% Add description of the optimization procedure.

As a demonstration, Fig. \ref{fig:loss_compare} shows the relative error of
original and optimized waveform at different frequencies. In this optimization
procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa}
as training waveforms to fit waveform coefficients.   
We can see the error of optimized waveform is lower than that of the original waveform for most of the domain.
In particular, the error in the early inspiral region is decreased by half while other regions also show good improvement in accuracy.

We can generalize the fitting procedure described above to a collection of waveforms.
In that case, the loss function is defined as the average of the mismatch of
individual waveforms, given by
\begin{equation}
	\mathcal{L}=\frac{1}{N}\sum_{i=1}^{N}\mathcal{M}_i,
    \label{eq:multi_loss}
\end{equation}
where $\mathcal{M}_i$ is the mismatch of individual training waveforms and $N$
is the total number of training waveforms used in optimization. This
optimization is more difficult since we are applying the same set of
coefficients to waveforms with different intrinsic parameters, such as mass
ratio and spins. From Eq. \ref{eq:multi_loss}, we can see the averaging between
waveforms with different intrinsic parameters implies there are trade-offs in
performance in different regions of the intrinsic parameter space.

@@

Forecasting the sensitivity of future experiments is an essential task in GW science. \kw{LOL I won't say it is essential.}
Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters can be measured.
The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
Although in practice this assumption is often violated in real GW datasets, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.

Computing the Fisher matrix requires one to evaluate derivatives of the likelihood which in turn involves derivatives of the waveform model and detector projection functions.
AD is therefore perfectly suited for computing Fisher matrices accurately and efficiently. 
Forecasting with Fisher matrices for third generation detectors has already been extensively explored in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
Here we purely want to illustrate the simplicity and speed of AD for forecasting rather than providing new physics insights.
We therefore consider a simple, three detector setup corresponding to the two LIGO detectors in addition to Virgo.

>>>>

\definecolor{dfm}{HTML}{6C71C4}
\newcommand{\dfm}[1]{{\color{dfm}[DFM: #1]}}
\newcommand{\amc}[1]{{\color{red}[AC: #1]}}

@@

\newcommand{\JHU}{\affiliation{William H. Miller III Department of Physics and Astronomy, Johns Hopkins University, Baltimore, Maryland 21218, USA}} 
\newcommand{\OKC}{\affiliation{The Oskar Klein Centre, Department of Physics, Stockholm University, AlbaNova, SE-106 91 Stockholm, Sweden}} 
\newcommand{\NORDITA}{\affiliation{Nordic Institute for Theoretical Physics (NORDITA), 106 91 Stockholm, Sweden}}
\newcommand{\Ciela}{\affiliation{Ciela -- Computation and Astrophysical Data Analysis Institute, Montréal, Quebec, Canada}}
\newcommand{\UdeM}{\affiliation{Département de Physique, Université de Montréal, 1375 Avenue Thérèse-Lavoie-Roux, Montréal, QC H2V 0B3, Canada}} 
\newcommand{\Mila}{\affiliation{Mila -- Quebec AI Institute, 6666 St-Urbain, \#200, Montreal, QC, H2S 3H1}} 
\newcommand{\CGP}{\affiliation{Center for Gravitational Physics, University of Texas at Austin, Austin, TX 78712, USA}}
\newcommand{\CCA}{\affiliation{Center for Computational Astrophysics, Flatiron Institute, New York, NY 10010, USA}}
\newcommand{\CUHK}{\affiliation{Department of Physics, The Chinese University of Hong Kong, Shatin, N.T., Hong Kong}}

@@

% Abstract with filler text
\begin{abstract}
    Since their discovery in 2015, gravitational waves (GWs) have developed into a unique observational probe used throughout fundamental and astro-phyics.
    As more detections continue, data analysis tasks are becoming increasingly computationally demanding.
    \kw{I would cut everything before out.}
    Here, we propose the use of automatic differentiation through the programming framework \jax for accelerating a variety of these analysis tasks.
    Firstly, we demonstrate that complete waveforms which cover the inspiral, merger, and ringdown of binary black holes (i.e.~IMRPhenomD) can be written in \jax and demonstrate that the serial evaluation speed of the waveform (and its derivative) is similar to the \lalsuite implementation in \CC.
    Moreover, \jax allows for GPU-accelerated waveform calls which can be over five orders of magnitude faster than serial evaluation on a CPU.
    We then focus on three applications where efficient and differentiable waveforms are essential.
    Firstly, we demonstrate how stochastic gradient descent can be used to optimize the $\sim 200$ coefficients that are used to calibrate the waveform model. 
    In particular, we demonstrate that the typical \textit{match} with numerical relatively waveforms can be improved by more than 10\% without any additional overhead.
    Secondly, we show that Fisher forecasting calculations can be sped up by $\sim 100\times$ (on a CPU) with no loss in accuracy.
    This increased speed makes population forecasting substantially simpler.
    % , as discussed extensively in \citet{Iacovelli:2022bbs, Iacovelli:2022mbg}.
    Finally, we show that gradient-based samplers like Hamiltonian Monte Carlo lead to significantly reduced autocorrelation values when compared to traditional Monte Carlo methods.
    Since differentiable waveforms appear to have substantial advantages for a variety of tasks throughout GW science, we propose that waveform developers use \jax to build new waveforms moving forward.
    Our waveform code, \ripple, can be found at \href{https://github.com/tedwards2412/ripple}{github.com/tedwards2412/ripple}, and will continue to be updated with new waveforms as they are implemented.

@@

Derivatives are ubiquitously useful throughout data analysis tasks.
% For instance, during PE, gradients are used to improve performance of the samplers 
% % \dfm{This isn't correct. In HMC the gradients don't have the behavior. I'd just say something like ``gradients are used to improve performance''} 
% (e.g.~in Hamiltonian Monte Carlo~\citep{2017arXiv170102434B} or Gradient Descent~\citep{2016arXiv160904747R}).
For instance, during PE, derivative information can be used to guide an optimizer towards higher-likelihood values (e.g.\ using gradient descent~\citep{2016arXiv160904747R}) or allow a sampler to rapidly explore parameter space (e.g.\ using Hamiltonian Monte Carlo~\citep{2011hmcm.book..113N,2017arXiv170102434B}).
Gradients are particularly valuable for high dimensional spaces.
Unfortunately, in the field of GW data analysis, analytic derivatives of the necessary quantities (such as the likelihood) have historically been difficult to obtain~\citep{Keppel:2013kia}.
\kw{This needs some rewording. For example, I think in IMRPhenomD they do have the analytical derivatives coded up, at least for a part of the waveforms. Instead of difficult to have the derivatives, I would rather mention it is clumpsy to get the derivatives, especially given the model complexity is growing.}
Numerical derivatives also suffer from accuracy issues stemming from rounding or truncation errors.
However, recent progress in automatic differentiation (AD) has shown promise in allowing general, fast derivative calculations for gravitational waveforms, with applications to constructing template banks~\citep{Coogan:2022qxs} and computing the Fisher information for forecasting~\citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.

Automatic differentiation is a family of methods used to compute machine-precision derivatives with little computational overhead. 
AD's recent ascendance is primarily driven by its use in machine learning, particularly for derivative computations of neural networks which use gradient descent during training.
The core idea of AD is that any mathematical function can be broken down into a small set of basic operations, each with a known differentiation rule.\footnote{
    Of course, non-differentiable functions exist and care must be taken when treating these special cases.
    }
The full derivative can then be constructed using the chain rule.
There are now a variety of AD implementations, most notably in deep learning frameworks such as \texttt{pytorch}~\citep{pytorch} and \texttt{tensorflow}~\citep{tensorflow2015-whitepaper}.
More general frameworks exist in \texttt{julia}~\citep{zygote, forwarddiff}, although \texttt{julia}'s limited use in GW analysis software precludes its general use.
Here we make use of \jax~\citep{jax2018github} due to its easy integration with \texttt{python} libraries, seamless support for running code on different hardware accelerators (e.g.\ graphical processing units) and its just-in-time (JIT) compiler, which can substantially accelerate code.

@@

In this paper we argue that differentiable waveforms will be a vital component for the future of GW data analysis.
\AZ{I prefer to describe what is rather than what could be; can we say what elements exist, e.g.~
we present \ripple... which includes \jax implementations of GW antenna patterns, the IMRPhenomD waveform, and...}
In addition, we present \ripple, a small GW \texttt{python} package which can serve as a repository for differentiable implementations of some of the main waveforms currently used in LIGO and Virgo analyses. 
The remainder of this paper is structured as follows. 
In Sec.~\ref{sec:waveforms} we discuss the differentiable IMRPhenomD waveform implemented in \ripple and perform some benchmarks to demonstrate its speed and accuracy. 
In Sec.~\ref{sec:applications} we discuss three distinct applications using differentiable waveforms. 
Firstly, we illustrate how the fit coefficients that form part of the IMRPhenom waveform models could be improved by high dimensional fitting enabled by a differentiable waveform. 
Secondly, we implement differentiable detector response functions and show that the speed of Fisher matrix calculations can be substantially accelerated using AD.
Finally, we run an illustrative injection example using Hamiltonian Monte Carlo to demonstrate that the autocorrelation of derivative samplers is substantially smaller than that of traditional MCMCs.
Practically, this translates into much faster PE with no sacrifice in accuracy.
The associated code can be found at \ripple~\citep{ripple}.

@@

A variety of waveform families have been developed to accurately model the GW emission from COs~\citep{Schmidtreview}. 
When the COs are relatively well separated, the dynamics of the system can be well approximated by a post-Newtonian expansion.
However, close to merger, numerical relatively (NR) simulations are required to accurately model the binary.
Unfortunately, these numerical simulations are computationally expensive and cannot be run in conjunction with data analysis.
Approximate, phenomenological waveforms have therefore been constructed to enable relatively fast waveform generation at sufficient accuracy.

As mentioned in the previous section, the three major waveform families that have been developed to date are:  EOB,  NRsurrogate, and IMRPhenom.
Note that unlike NRsurrogate, both the EOB and IMRPhenom waveforms must be calibrated to numerical relativity waveforms.
EOB waveforms require one to model the binary system using a Hamiltonian and are typically slow to evaluate, whereas IMRPhenom waveforms are constructed with simple closed-form expressions.
IMRPhenom waveforms are therefore ideally-suited for AD, especially using \jax. 
In this paper we focus on the aligned-spin, circular-orbit model IMRPhenomD~\citep{Husa:2015iqa, Khan:2015jqa}.

@@

For maximum utility, a waveform needs to be fast to evaluate.
Fortunately, the IMRPhenom waveforms are constructed from simple closed-form expressions which are computationally efficient.
Since the \lalsuite implementation of IMRPhenomD is written in \texttt{C}, the evaluation of the waveform itself is faster than \ripple. \dfm{I'd soften this statement a little bit. In my own work, I often find that JAX implementations are just as fast or faster than C implementations, so I wouldn't state it that it's obvious that lalsuite should be faster! Where are these benchmarks implemented? Would it be easy for me to try them out?}
Benchmarking on a MacBook Pro with an M1 Max Apple Silicon processor, we find that a single waveform evaluation takes $\sim 2-5\,\mathrm{ms}$ for \lalsuite (interfaced with python) and $\sim 5-8\,\mathrm{ms}$ for \ripple.\footnote{
    For this benchmark we used $16\,\mathrm{Hz}$ and $1024\,\mathrm{Hz}$ for the lowest and highest frequencies respectively. 
    In addition, we used a frequency spacing of $0.0125\,\mathrm{Hz}$. 
    We performed this benchmark by evaluating the waveform $5000$ times and taking the average evaluation time.
}
Although slower on a CPU benchmark, a key advantage of using \jax is its native ability to run on a GPU.
Waveform calls can therefore be evaluated in parallel, massively improving performance.
Performing the same benchmark as above on a Google Colab notebook with a freely-provided NVIDIA Tesla T4 GPU (far from the current state-of-the-art), we find that on average waveform evaluations take $\sim \sout{0.00005} 5 \times 10^{-5}\mathrm{ms}$, nearly five orders of magnitude faster.
\kw{We should add some more detail related to this comparison. We should be clear that each individual waveform sample evaluation doesn't get us the performance by putting it on GPU (Jitting helps with that, not the GPU part). The GPU comes into play by allowing parallel evaluation of waveform sample, so for a tiny amount of time/frequency samples, we should expect GPU to lose to a CPU, but on any reasonable amount of time/frequency samples, GPU win a lot.}
\amc{mention \texttt{vmap} and/or \texttt{pmap} -- super trivial to parallelize! More work in C.}
Generalizing \lalsuite waveforms to run on a GPU would be a significant undertaking.

The primary argument of this paper is that waveform derivatives will also be highly valuable to data analysis tasks. 
AD provides two big advantages when it comes to evaluating derivatives compared to numerical differentiation.
First, the accuracy of derivatives from AD are significantly more stable than finite-difference methods.
In particular, finite differences suffer from both rounding and truncation errors, meaning that the user is required to \textit{tune} the width over which the difference is taken.
On the other hand, AD produces machine-precision derivatives with no tuning.
Second, AD scales favorably with the dimensionality of the function.
In particular, for every input dimension, $D$, added one would need to evaluate the function at least $2D$ times to calculate finite difference derivatives for all input parameters.
For reverse-mode AD, one only needs two function calls to evaluate the derivative of all input parameters, regardless of dimension.\footnote{
    Note that although the number of function calls is small, reverse-mode AD does add memory overhead.
    We've not found this to be limiting in any of the situations tested so far.
}
\kw{Maybe a bit more background on AD, the way we phrase it here sounds a bit too magical too me now.}
Since the parameter space of GWs in GR has $\mathcal{O}(10)$ dimensions, the speed of derivative evaluation is less crucial than the stability.
However, this might change for waveforms in beyond-GR models, \AZ{waveforms involving equation of state parameters for neutron stars, and models which account for calibration and waveform uncertainties.
In these cases} many more parameters \AZ{can be} added.

@@

\subsection{Fine-Tuning Waveform Coefficients}
\label{subsec:coeffs}

@@

Having an accurate waveform model is essential for many data analysis tasks in
GW, such as searching for signals and estimating source parameters given data.
While waveforms generated using numerical relativity simulations are \sout{commonly
regarded as the most accurate,} 
\AZ{in principle the highest fidelity signal model,}
they are too computationally expensive to be used
in any practical data analysis tasks. To create waveform models that can be used
in data analysis, there are a number of groups creating ``approximants'' of the
full numerical relativity waveforms, which are essentially simpler ansatzes that
\sout{are only approximations but} can be evaluated much faster \AZ{and in regions not
covered by numerical simulations, such as extremes of parameter space or earlier 
phases of the binary inspiral}. 
\amc{redundant -- remove or merge with earlier material.}

@@

Previously in the construction of IMRPhenomD \citep{Khan:2015jqa}, waveform
coefficients were fitted in segments with amplitude and phase separated.
Furthermore, IMRPhenomD is divided in three fitting segments, inspiral, merger and ringdown.
Each of these segments has their own set of fitting coefficients.
After obtaining the fitting coefficient for individual segments, they are then ``stitched'' together such that they are continuous in the first derivative.
The process of stitching introduces some additional
inaccuracy in the waveform model, as the connections would affect the originally
fitted segments.

The coefficients of the current IMRPhenomD implementation are tuned in subsets
of parameters instead of all together. This means the tuning process ignores the
correlation between different subsets of parameters, so the best-fit
solution obtained by fitting subsets of parameters may not be the global
optimum compared to fitting all coefficients at once. By jointly
fitting all coefficients, the correlation in fitting individual and connecting
different segments can be accounted for during the optimization. Therefore, there is
potential improvement of the accuracy by fitting all coefficients at once.

@@

The first step in fitting all coefficients is to define a loss function that
measures the goodness-of-fit of the optimization procedure, which we choose it
to be the mismatch between the NR waveform and the approximant waveform:
\begin{equation}
    \label{eq:mismatch}
	\mathcal{M}(\lambda)=1-\mathrm{m}(h_{\mathrm{IMR}}(\lambda)|h_{\mathrm{NR}}(\lambda)),
\end{equation}
where $\lambda$ is a vector of the fitting coefficients, $h_{\mathrm{IMR}}$ is
the waveform generated by the IMRPhenomD and $h_{\mathrm{NR}}$ is the waveform
generated by the numerical relativity simulation. 
Given the loss function, we use gradient descent to find the optimal fitting
coefficients.
\begin{equation}
	\lambda\leftarrow\lambda-\alpha\nabla\mathcal{M}, 
\end{equation}
where $\alpha$ is the learning rate. We set $\alpha$ to be a very small number, i.e. $10^{-6}$, since the initial guess already lies close to the desired optimum. We terminate the program after 30000 steps, in which $\lambda$ has converged to the optimum.  \dfm{We can't possibly know that it has converged to ``the optimum''. I'd sat just leave it at ``We terminate the program after 30000 steps.''}
\amc{say something to explain why this is working. I.e.: does the loss initially drop and then plateau?}
% Add description of the optimization procedure.

As a demonstration, Fig. \ref{fig:loss_compare} shows the relative error of
original and optimized waveform at different frequencies. In this optimization
procedure, we took NR waveforms $h_{\mathrm{NR}}$ stated in \citep{Khan:2015jqa}
as training waveforms to fit waveform coefficients. \amc{how many are there?}
We can see the error of optimized waveform is lower than that of the original waveform for most of the domain.
In particular, the error in the early inspiral region is decreased by half while other regions also show good improvement in accuracy. \amc{so this is a waveform in the training set? And I don't follow exactly what you did. Is this just fitting a single waveform?}

We can generalize the fitting procedure described above to a collection of waveforms.
In that case, the loss function is defined as the average of the mismatch of
individual waveforms, given by
\begin{equation}
	\mathcal{L}=\frac{1}{N}\sum_{i=1}^{N}\mathcal{M}_i,
    \label{eq:multi_loss}
\end{equation}
where $\mathcal{M}_i$ is the mismatch of individual training waveforms and $N$
is the total number of training waveforms used in optimization. This
optimization is more difficult since we are applying the same set of
coefficients to waveforms with different intrinsic parameters, such as mass
ratio and spins. From Eq. \ref{eq:multi_loss}, we can see the averaging between
waveforms with different intrinsic parameters implies there are trade-offs in
performance in different regions of the intrinsic parameter space.
\amc{this means your fit generally will depend on how the training waveforms are distributed over parameter space.}

@@

Forecasting the sensitivity of future experiments is an essential task in GW science. \kw{LOL I won't say it is essential.}
Due to its theoretical simplicity and evaluation speed, the Fisher matrix formalism is commonly deployed to estimate how well a binary system's parameters can be measured.
The Fisher matrix approach is built around the assumption of a Gaussian likelihood.
Although in practice this assumption is often violated for realistic detector noise, the results obtained using a Fisher analysis can provide quick and useful diagnostics in evaluating sensitivities for a variety of models and detector configurations.

Computing the Fisher matrix requires one to evaluate derivatives of the likelihood, which in turn involves derivatives of the waveform model and detector projection functions.
AD is therefore perfectly suited for computing Fisher matrices accurately and efficiently. 
Forecasting with Fisher matrices for third generation detectors has already been extensively explored in \citep{Iacovelli:2022bbs, Iacovelli:2022mbg}.
Here we purely want to illustrate the simplicity and speed of AD for forecasting rather than providing new physics insights.
We therefore consider a simple, three-detector setup corresponding to the two LIGO detectors in addition to Virgo.

The Fisher information matrix for a single detector is typically given by \amc{I'm guessing in practice you compute a Hessian, so maybe write that out?}
\begin{equation}
    \label{eq:fisher}
    \mathcal{I}^{k}_{ij} = (\partial_i h^k | \partial_j h^k) \, ,
\end{equation}
where $k$ indicates the detector, $\partial_i = \partial/\partial \theta_i$, and $h$ is the strain measured by the detector which is given by,
\begin{equation}
    h^k(\theta) = F_+(\phi) h^k_{+}(\Xi) + F_\times(\phi) h^k_{\times}(\Xi) \, .
\end{equation}
Note that here we have separated out the extrinsic ($\phi$) and intrinsic ($\Xi$) variables as well as introducing the detector projection functions for the plus and cross polarizations as $F^k_+$ and $F^k_\times$ respectively.
Since we are considering a three detector setup we simply add the Fisher matrices from the individual detectors to get the combined Fisher matrix:
\begin{equation}
    \mathcal{I}_{ij} =  \mathcal{I}^{\mathrm{Hanford}}_{ij} + \mathcal{I}^{\mathrm{Livingston}}_{ij} + \mathcal{I}^{\mathrm{Virgo}}_{ij}   \, .
\end{equation}
Finally, we invert the Fisher matrix to calculate the covariance matrix,
\AZ{which proves forecasted measurement errors and parameter covariances for a signal with parameters $\theta$ observed by the given detector network}.
\AZ{Again we need to state and cite the PSDs we use for the three detectors.}
